---
title: "Storm Data Analyses"
author: "Jordan A. Kempker, MD, MSc"
date: "January 25, 2016"
output: html_document
---

# Synopsis
The United States (US) National Oceanic and Atmospheric Administration (NOAA) has tracked and documented major weather events impacting the US since 1950.  The database captures the time and location of severe weather events as well as estiamtes of their human impact.  

The following analyses will examine this database.  Information regarding how the database can be found at the following links:
> National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
> National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

# Data Processing
The data processing and analysis will need the following R packages.
```{r}
library(dplyr)
library(ggplot2)
```

1. Data download and load into R.
    + After manually setting your working directory in R with setwd(), the following code will download the data into your working directory and then load into an R object, named **storm**.
    ```{r, cache = TRUE}
    if (!file.exists("./StormData.csv.bz2")){
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url,destfile = "./StormData.csv.bz2", method="auto")
    }
    storm <- read.csv("./StormData.csv.bz2", stringsAsFactors = TRUE)
```
2. To get a sense of the data we will run a str() and head().
```{r}
str(storm)
head(storm)
```
3. Since we are doing an analysis by storm type, we will transform the **EVTYPE** into a factor and group data by this factor variable.
```{r}
storm$EVTYPE <- as.factor(storm$EVTYPE)
storm <- group_by(storm, EVTYPE)
```
4. 
# Results